Model Finding for Recursive Functions in SMT\thanks{ This research is partially supported by the Inria technological development action ``Contre-exemples Utilisables par Isabelle et Coq'' (CUIC). }

### 2015-05-07 tex/main.tex
Changes to abstract, minor.

\begin{abstract}
SMT solvers have recently been extended with techniques for finding models
in presence of (universally) quantified formulas.
However, these generally cannot cope with universal quantifiers ranging over
infinite types, such as integers and algebraic datatypes. This paper introduces
a translation that eliminates infinite quantifiers in axioms specifying a large
class of recursive functions, including terminating and
productive functions. An empirical evaluation confirms that the approach works
well on benchmarks from two sources. The translation is implemented as a
preprocessor in the solver CVC4.
\end{abstract}

@@

\begin{definition}\rm
Given a set of function definitions 
$\DDD = \{ \forallf{\con{f}_1} \vec x.\; \varphi_1, \ldots, \forallf{\con{f}_n} \vec x.\; \varphi_n \}$, 
a ground formula $\varphi$ 
%\ct{I suppose $\varphi$ is meant to be ground.} yes --ajr
is \emph{closed under function expansion with respect to $\DDD$} if 
\[
 \varphi \,\models_T\, \bigwedge\nolimits_{\,i=1}^{n} \{ \varphi_i[ \vec t / \vec x ] \mid \con{f}_i( \vec t ) \in \terms( \varphi ) \} .
\]
The set $\DDD$ is \emph{admissible} if for every $T$-satisfiable formula 
$\varphi$ closed under function expansion with respect to $\DDD$,
the formula
$\varphi \wedge (\forallf{\con{f}_1} \vec x.\; \varphi_1) \wedge \cdots \wedge (\forallf{\con{f}_n} \vec x.\; \varphi_n)$ is also $T$-satisfiable.
\ct{I}
\end{definition}

Admissibility is a semantic criterion that must be satisfied for each function
definition before applying the translation described in
Section~\ref{sec:encoding}. It is interesting it connect it with the more
standard notion of \emph{well-founded} function definitions, often called
``terminating'' definitions in a slight abuse of terminology. In such
definitions, all recursive calls are decreasing with respect to a well-founded
relation $r$, which must be supplied by the user or inferred automatically
using a termination prover. This ensures that the function is uniquely defined
at all points.

@@

Another interesting fragment is the class of \emph{productive} corecursive
functions. Corecursive functions are functions to a coalgebraic datatype.
These functions can call themselves forever, without their being inconsistent.
Productive corecursive functions are functions that progressively reveal parts
of their (potentially infinite)
output \cite{turner-1995,abbott-et-al-2005,mcbride-productive}.
For example, given a type of infinite streams constructed by $\con{scons}$,
the definition
$\{\forallf{\const{e}} n.\;\, \const{e}(n) \teq \con{scons}(n,\; \const{e}(n + 1))\}$
falls within this fragment, since each call to $\const{e}$ produces one
constructor before entering the nested call. Like terminating recursion,
productive corecursion uniquely specifies a function. The following theorem
connects these notion with admissibility.

\begin{theorem}
Let 
\end{theorem}

@@

The existential $\exists b : {\fargtype{\con{s}}}.\;\, (\farg{\con{s}}( b ) \teq \con{c}-1)$ above
ensures that whenever $\I$ satisfies the set $L_1 = L_0 \cup \{ \neg \con{c} \leq 0, \con{s}(\con{c}) \teq \con{c} + \con{s}( \con{c}-1 ) \}$,
$\I$ satisfies $\absconstraints( L_1 )$ as well.
Hence, a model of formula~(\ref{eq:ex-after}) that interprets $\con{c}$ as $n$ must also satisfy the formula $\psi$:
%
\begin{equation} \label{eq:fun-closure}
\begin{split}
& \con{s}( \con{c} ) > 100 \wedge %\displaystyle
  \bigwedge\nolimits_{\vthinspace i=0}^{\!n-1} \bigl( \neg (\con{c}-i \leq 0) \wedge \con{s}( \con{c}-i ) \teq (\con{c}-i) + \con{s}( (\con{c}-i)-1 ) \bigr) \\[-\jot]
{\wedge}\; & \con{c}-n \leq 0 \wedge \con{s}( \con{c}-n ) \teq 0
\end{split}
\end{equation}
%
This formula is closed under function expansion,
since it contains the $\con{s}$ applications corresponding to
$\con{s}( \con{c}-i )$ for $i = 0, \ldots, n$. The formula also
entails $\varphi_\con{s}[\con{c}-i/x]$ for $i = 0, \ldots, n$.
Since $\varphi_\con{s}$ is an admissible definition,
there must exist a $\Sigma$-interpretation satisfying $\psi \wedge \forallf{\con{s}} x.\;\, \varphi_\con{s}$
which entails formula~(\ref{eq:ex-before}).\xend
\end{example}

>>>>

\begin{abstract}
SMT solvers have recently been extended with techniques for finding models
in presence of (universally) quantified formulas.
These techniques are generally limited to universally quantified formulas in some restricted fragment.
This paper introduces a translation which reduces axioms specifying a large
class of recursive functions, including terminating, tail-recursive, and
productive functions to universally quantified formulas for which known SMT solving techniques are applicable.
An empirical evaluation confirms that the approach improves 
the performance of existing solvers on benchmarks from two sources. 
The translation is implemented as a preprocessor in the solver CVC4.
\end{abstract}

@@

\begin{definition}\rm
Given a set of function definitions 
$\DDD = \{ \forallf{\con{f}_1} \vec x.\; \varphi_1, \ldots, \forallf{\con{f}_n} \vec x.\; \varphi_n \}$, 
a ground formula $\varphi$ 
%\ct{I suppose $\varphi$ is meant to be ground.} yes --ajr
is \emph{closed under function expansion with respect to $\DDD$} if 
\[
 \varphi \,\models_T\, \bigwedge\nolimits_{\,i=1}^{n} \{ \varphi_i[ \vec t / \vec x ] \mid \con{f}_i( \vec t ) \in \terms( \varphi ) \} .
\]
The set $\DDD$ is \emph{admissible} if for every $T$-satisfiable formula 
$\varphi$ closed under function expansion with respect to $\DDD$,
the formula
$\varphi \wedge (\forallf{\con{f}_1} \vec x.\; \varphi_1) \wedge \cdots \wedge (\forallf{\con{f}_n} \vec x.\; \varphi_n)$ is also $T$-satisfiable.
%\ct{I}
\end{definition}

Admissibility is a semantic criterion that must be satisfied for each function
definition before applying the translation described in
Section~\ref{sec:encoding}. It is interesting it connect it with the more
standard notion of \emph{well-founded} function definitions, often called
\emph{terminating} definitions in a slight abuse of terminology. In such
definitions, all recursive calls are decreasing with respect to a well-founded
relation $r$, which must be supplied by the user or inferred automatically
using a termination prover. This ensures that the function is uniquely defined
at all points.

@@

Another interesting fragment is the class of \emph{productive} corecursive
functions. Corecursive functions are functions to a coalgebraic datatype.
These functions can call themselves forever, without their being inconsistent.
Productive corecursive functions are functions that progressively reveal parts
of their (potentially infinite)
output \cite{turner-1995,abbott-et-al-2005,mcbride-productive}.
For example, given a type of infinite streams constructed by $\con{scons}$,
the definition
$\{\forallf{\const{e}} n.\;\, \const{e}(n) \teq \con{scons}(n,\; \const{e}(n + 1))\}$
falls within this fragment, since each call to $\const{e}$ produces one
constructor before entering the nested call. Like terminating recursion,
productive corecursion totally specify the functions it defines. The
following theorem connects these notion with admissibility.

\begin{theorem}
Let
$\DDD = \{ \forallf{\con{f}_1} \vec x.\; \varphi_1, \ldots, \forallf{\con{f}_n} \vec x.\; \varphi_n \}$
be a set of function definitions such that
%
Then the set $\DDD$ is admissible.
\end{theorem}
\begin{proofsketch}
\qed
\end{proofsketch}

@@

The existential $\exists b : {\fargtype{\con{s}}}.\;\, (\farg{\con{s}}( b ) \teq \con{c}-1)$ above
ensures that whenever $\I$ satisfies the set $L_1 = L_0 \cup \{ \neg \con{c} \leq 0, \con{s}(\con{c}) \teq \con{c} + \con{s}( \con{c}-1 ) \}$,
$\I$ satisfies $\absconstraints( L_1 )$ as well.
Hence, by repeated applications of this reasoning, it follows that
a model of formula~(\ref{eq:ex-after}) that interprets $\con{c}$ as $n$ must also satisfy the formula $\psi$:
%
\begin{equation} \label{eq:fun-closure}
\begin{split}
& \con{s}( \con{c} ) > 100 \wedge %\displaystyle
  \bigwedge\nolimits_{\vthinspace i=0}^{\!n-1} \bigl( \neg (\con{c}-i \leq 0) \wedge \con{s}( \con{c}-i ) \teq (\con{c}-i) + \con{s}( (\con{c}-i)-1 ) \bigr) \\[-\jot]
{\wedge}\; & \con{c}-n \leq 0 \wedge \con{s}( \con{c}-n ) \teq 0
\end{split}
\end{equation}
%
This formula is closed under function expansion,
since it contains the $\con{s}$ applications corresponding to
$\con{s}( \con{c}-i )$ for $i = 0, \ldots, n$. The formula also
entails $\varphi_\con{s}[\con{c}-i/x]$ for $i = 0, \ldots, n$.
Since $\varphi_\con{s}$ is an admissible definition,
there must exist a $\Sigma$-interpretation satisfying $\psi \wedge \forallf{\con{s}} x.\;\, \varphi_\con{s}$
which entails formula~(\ref{eq:ex-before}).\xend
\end{example}
